Implementation and assessment of methods for projecting forest road network development

### 2023-04-19 analysis/paper/paper.Rmd
writing in progress

---
title: "Assessment and implementation of resource road simulation methods for projection of cumulative effects of disturbance on wildlife"
author:
  - name: Josie Hughes
    email: josie.hughes@ec.gc.ca
    institute: [ECCC]
    correspondence: true
    equal_contributor: "yes"
    orcid_id: 0000-0001-7875-9015
  - name: Sarah Endicott
    email: sarah.endicott@ec.gc.ca
    institute: [ECCC]
    correspondence: false
    equal_contributor: "yes"
  - name: David Lapins
    email: david.lapins@pc.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Kyle Lochhead
    email: Kyle.Lochhead@gov.bc.ca
    institute: [FLNR]
    correspondence: false
  - name: Gregory Paradis
    email: gregory.paradis@ubc.ca
    institute: [UBC]
    correspondence: false

institute:
  - ECCC: Environment and Climate Change Canada 
  - FLNR: British Columbia Ministry of Forests 
  - UBC: University of British Columbia 
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      number_sections: no
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: Roads.bib
csl: "../templates/journal-of-forest-ecology-and-management.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
editor_options: 
  chunk_output_type: inline
---

@@

@@

Leston et al. (2020) projected forest road network development using Patchworks [@spatial_planning_systems_patchworks_nodate], then added fire. This example highlights that roads are also important for harvest planning, and that road network planning methods [@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997] and tools exist [@remsoft_remsoft_2019; @spatial_planning_systems_patchworks_nodate; @epstein_combinatorial_2006]. However, algorithms contained in proprietary closed-source software are difficult to assess or fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017]. Methods intended to support tactical and operational planning also require detailed parameterization with inputs that may not be publicly available, and employ algorithms that are too complex for application at regional, provincial or national scales. More promising options for integrating forest road network projections into large scale stochastic projections of forest change include an iterative least cost path method developed by Hardy et al [-@hardy_forest_2019; -@hardy_landis-ii_2023], and least cost path and minimum spanning tree algorithms used to project roads and reconstruct development history in British Columbia [@lochhead_demo_2018; @lochhead_linking_2022]. The former method is available as QGIS and Landis-II extensions, and the latter is embedded in a larger open source modelling project that is specific to British Columbia [@muhly_castor_nodate]. Hardy's tool is designed for integration with other components of the Landis-II modelling framework [@scheller_design_2007; @labadie_global_2023], but none of the existing implementations can be easily integrated into the full variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g. @barros_empowering_2023; @daniel_state-and-transition_2016].

@@

<!-- Our goal is provide simple, accessible, and usable forest road network projections that can easily be applied in the context of large-scale landscapes and wildlife impact simulations. Our model will be openly accessible and uses minimal input parameters (an existing road network, harvest cutblocks, and cost surface) which makes it easily transferable to many forest resource dominated landscapes. Here, we evaluate its accuracy at simulating forest road development across a large landscape by recreating the network starting from approximately 1990 and comparing it to the present day observed forest road network. In order to compare the projected and observed road networks, we used various metrics that are similar to those used in wildlife impact studies to compare the projected and observed landscapes both spatially and aspatially. -->

@@

We chose the Revelstoke timber supply area (TSA 27), in British Columbia, Canada as an assessment area because it is a mountainous areas with approximately 55% of non-forested land where road building is difficult or impossible (e.g. alpine, lakes, swamp, brush, rock) [@hachey_revelstoke_2010]. Thus, the area provides opportunities to observe the responses of road projection algorithms to steep complex terrain. Forestry is the dominant resource extraction activity in this region and much of the road network was built to access forest harvest cutblocks (Figure \@ref(fig:study-area)), which allows projection of the forest road network from historic forest harvest data.  Revelstoke is in the interior-wet belt of the province where forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%) [@ministry_of_forests_revelstoke_2021]. The area is home to wildlife of concern including including grizzly bears and Southern Mountain woodland caribou [@hachey_revelstoke_2010]. Eight herds of Southern Mountain caribou have been extirpated or likely extirpated, and remaining herds are considered Endangered by COSEWIC and listed as threatened under the Canadian Species at Risk Act [@palm_long_2020]. Mountain caribou avoid roads leading to a loss of available habitat [@polfus_identifying_2011; @apps_factors_2006], suffer increased mortality near roads [@apps_spatial_2013], and decline in abundance as the footprint of roads and cutblocks increases [@lochhead_linking_2022]. Roads also alter grizzly habitat use, home range selection, movements, population fragmentation, survival, reproductive rates, population density, trends, and conservation status [@proctor_effects_2020].

@@

Road network data was obtained by combining the digital road atlas (DRA) [@data_bc_digital_2021] and forest tenure road [@british_columbia_ministry_of_forests_lands_natural_resource_operations_and_rural_development__flnrord_forest_2021] data sets for BC. To avoid situations where the same road was included in both data sets with slightly different coordinates, we removed all roads from the DRA data set that were within 50 m of a road in the forest tenure road data set. The existing road network, used as the starting point for projections, was created by filtering out forest tenure roads with an award date greater than 1990. All roads from the DRA data set were assumed to be built before 1990. Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset [@government_of_british_columbia_harvested_2021], compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes. The dataset does not exclude non-harvested 'reserved' areas, and satellite image change detection may not always represent the entire cutblock area [@government_of_british_columbia_harvested_2021]. Forest harvest cutblocks with harvest years greater than 1990 were filtered and used as targets for the projections.

We developed a simple model of road building cost informed by slope and major barriers (Table 1). Following the Interior Appraisal Manual of BC, we assume the base cost per kilometer of road is a linear function of slope [@government_of_british_columbia_interior_2021], and that road construction across major water bodies is prohibitively expensive (Table \@ref(tab:cost-tbl)). Existing roads are assigned a cost of 0. The cost raster consists of 1,703,936 pixels with a resolution of 1 ha.

@@

All data sets were accessed using the `bcdata` R package and filtered to the extent of the Revelstoke Timber Supply Area [@teucher_bcdata_2021]. These data sets are continuously updated so we have stored the versions used in our analysis in an Open Science Framework repository <https://osf.io/8vmqh/>. We were not granted permission to share the DRA data set so it must be accessed from <https://catalogue.data.gov.bc.ca/dataset/bb060417-b6e6-4548-b837-f9060d94743e>. Scripts used for data acquisition and analysis associated with this paper are available on GitHub <https://github.com/LandSciTech/RoadPaper>.

@@

We focus on methods that link multiple target locations to an existing road network for strategic purposes, described by Dean (1997) as the multiple target access problem (MTAP), where the objective is to find the minimum-cost locations for new roadways that link landings within harvest cutblocks (i.e., log decks) to the existing road network. We implemented four methods for solving the MTAP in the roads package, representing a gradient of computational complexity and operational detail; in this manuscript we focus on comparing the dynamic least cost path (DLCP) and minimum spanning tree approaches (MST) to an alternative iterative least cost path method developed by Hardy [-@hardy_forest_2019;-@hardy_landis-ii_2023] (Figure \@ref(fig:proj-meth-fig)).

The DLCP approach solves the MTAP by decomposing the problem into several smaller single target access problems (STAP). For each STAP, a least cost path from a single source to sink is solved using Dijkstra's shortest path algorithm [-@dijkstra_note_1959]. The algorithm is parameterized with a weighted graph wherein each node corresponds to a raster pixel representing one of three components of the potential forest road network: a landing, a point on an existing road, or an intermediate point that is a plausible location for road construction. Each node is connected with up to eight neighbours, and the cost of construction along edges is weighted by the distance to neighbouring nodes and the average of the construction cost at the source and destination nodes (Table \@ref(tab:cost-tbl)). The underlying grid pattern in these methods does not support curvature or switchbacks required in operational road design. The DLCP approach solves each STAP sequentially, starting with target that is nearest to an existing road. The existing road network and construction cost surface is updated with each new road segment before moving on to the next nearest target. The method was first developed by Anderson and Nelson (2004) who noted that the projected network is sensitive to the ordering of the STAP and recommended a heuristic for connecting the landings and finding optimal branch locations to deal with this sensitivity.

The MST approach extends the LCP approach by providing a heuristic for linking landings or evaluating road branching opportunities. Dean (1997) showed that heuristics with branch evaluation were able to reproduce 80% of the observed road locations in their study area. The first step in our MST approach is to estimate the least cost paths that link each of the landings and the least cost paths that link each landing to the existing road network. These results are used to create a weighted graph wherein nodes are either landings or points on the existing road network, and edge weights correspond to the total cost of the least cost path between pairs of nodes. The minimum spanning tree for this graph is solved using Kruskal's algorithm [-@kruskal_shortest_1956], which links every node in graph with a tree that minimizes edge weights and does not include loops. This method differs from the minimum spanning tree method used by Clark et al. (2000) who used Euclidean distances rather than least costs for edge weights.

The DLCP and MST methods connect target or landing locations within each cutblock to the existing road network, so an important first step is to select one or more landing locations in each block. We investigated the sensitivity of results to the pattern and density of landing locations by comparing results derived from the centroid of each cutblock to regular or randomly selected landing densities of 1 (low) and 10 (high) points per km^2^ (Figure \@ref(fig:proj-meth-fig)). We used default settings in the Hardy QGIS plugin except for skidding distance which was set to 200m. A skidding distance of 200 m corresponds to \~ 9 pts/km^2^. We compared two resolutions of the cost surface by aggregating the 1 ha resolution cost raster to a coarser 100 ha (1000 by 1000 m) resolution. In total, we compared 12 road projection method variants to a cutblocks only scenario with no projected roads (Table \@ref(tab:method-tbl)).

@@

TO DO: Sarah are these statements still correct? Please revise as necessary.

@@

TO DO: Sarah figure out why `r dlcp_time` and `r mst_time` are coming out as NA, and how to fix that.

@@

```{=html}

@@

<!-- TO CONSIDER: provide example script that would make it relatively easy for someone to assess the performance of a new road metric of their choice. They don't need to run the projections to do this - just summarize the outputs using a method of their choice. -->

<!--  What David had
Through assessing the performance of our model by comparing its spatial and aspatial accuracy with the observed forest road network following 1990, we are satisfied that the projections provide adequate results given their intended use, and that it is a useful tool for projecting forest road networks across a large landscape with minimal input parameters. However, for applications where the exact location of roads is important (like operational road engineering purposes), or for projecting impacts on wildlife that respond to roads at a very fine spatial scale then this is not an appropriate model. The model does, however, have potential to be a useful tool for predicting impacts on wildlife across large landscape simulations and could easily be incorporated as an element into more complex models that take into account other landscape phenomena such as wildfires, climate change, etc. Because this model does not take timescales into account, it could be a useful post-processing tool as an element of more complex models as it can be applied retroactively to landscape simulations.

Many of the major discrepancies between the projections and the observed network had to do with potential issues with the available data. Often major discrepancies were related to cutblocks that were likely incorrectly labelled (most often through LANDSAT change detection) in places where there weren't actually cutblocks. These discrepancies had a negative influence on the overall spatial and aspatial accuracy of the projections. In order to get more accurate road projections, one could simply remove the cutblocks derived from LANDSAT as these were often the cutblocks which caused issues, however, these cutblocks were left in as a major goal of this assessment was to create road projections while adjusting input parameters as little as possible. Considering how this model will be used to simulate forest roads moving into the future, it is unlikely that potential future cutblocks will be mislabelled as they were in our situation and this would not be an issue. *Another potential issue was with roads not being included in the data set that were seen in the satellite imagery.* (this is no longer identified in the results section since it wasn't a major discrepancy after fixing the road data source) This issue relates to results from the study by Poley et al (in review) that demonstrated how road network datasets are almost universally incomplete and that many low-use roads (such as forest resource roads) can be seen within recent satellite imagery but are not included in any readily available road datasets. Mapping of roads is often especially lacking in more rural, forested, and mountainous regions where roads are often unpaved (Prendes et al., 2017) with the Auditor General of British Columbia (2017) suggesting that 100 000 km of resource roads are missing from publicly-available road network maps (inspired by Adam Ford's doc). This means there were areas where the projections built a road, causing a discrepancy, but in reality there was actually an observed road there as seen in the satellite imagery. These discrepancies also negatively impacted the spatial and aspatial accuracy results when they shouldn't have and should be treated with caution. It is possible that some roads seen in the satellite images have since been closed which could be a reason for them being absent in the roads data sets. Again, this would not present an issue when predicting future impacts on landscapes, only when assessing the model accuracy in this context. Having said this, the projections were deemed most accurate in areas where the quality of the cutblock and road data was high and consistent with what was seen in recent satellite images.

Although this model works well in an area like Revelstoke where the road network is located in a region dominated by forestry roads, it does not work well in landscapes where roads are developed based on other targets outside of cutblocks (where the intent to build the road is not to achieve the least cost path; aka seismic lines). Fort Nelson TSA (TSA 8), for example, contains a large amount of non-forestry roads which are related to seismic lines, oil and gas extraction, etc (\@ref(fig:ftNelson)). Because of this, forestry roads make up a much smaller percentage of the overall road network. In landscapes like this, it does not make sense to use the forestry road projections as many roads are influenced by many different factors. However, other resource extraction activities could likely be incorporated into this model relatively easily. The same concept could be used where mining or oil locations could be used as target locations instead of cutblock areas.

It is also worth noting that although the model can project road networks fairly accurately across a large landscape, it may differ from the way in which roads are planned in real life. The model will simply build the roads in the least costly locations based on the cost surface, however often in the real world, roads must be planned around ownership and administrative boundaries. In BC, forestry roads are developed by forest licensees under a timber harvesting permit (Government of British Columbia, 2021c).

Certain projection methods were more accurate at projected some metrics compared to others. Similarly, certain metrics were sensitive to pixel size (road density, road presence) and others were not. Depending on what the goal of running the projections is, then the projection method and resolution may or may not play an important role. If you would like to predict road density as accurately as possible, then running the projections using as fine a resolution of cost surface as possible is the best option. If you're wanting to only predict roughly the location of the roads/save computing time and are more interested in a larger area, then a coarse resolution may suit these requirements perfectly fine. The tradeoffs of using a coarser resolution of cost surface may be less extreme when projecting roads across a flatter landscape compared to a mountainous landscape like was used for this paper since there would likely be less variability in cost across larger areas.

With caribou being one of the key considerations when managing some forestry areas (like in British Columbia TSAs, including Revelstoke TSA) (Government of British Columbia, 2021), the accuracy at which "roads" can project this forestry disturbance footprint metric could allow for this package to be a truly beneficial tool for assessing the impacts towards at-risk caribou populations when looking at various forest harvest scenarios. Because caribou are also considered an umbrella species within forests (Drever et al., 2019), projecting forestry impacts on their habitats may also provide valuable insight towards other key species as well. With the input of future cutblocks and the projections of forest road networks to these cutblocks, the results from the projection can provide a great outlook at the impacts of forestry scenarios on woodland caribou habitats using the forestry disturbance footprint metric. The distance to nearest road metric appears to be a very interesting metric to consider as well when looking specifically at larger areas. Using this metric to make projections at larger scales, such as at a TSA level, may provide very beneficial information about forestry impacts especially in a wildlife context when assessing the influence of forestry on various species with different tolerances to disturbance.

Further improvements could be made that would increase the accuracy of the model. The cost surface used for this paper was developed at a provincial scale, however, a higher resolution cost surface with more detail could further improve the accuracy of the projection results. A finer resolution cost surface with more detail will ultimately create a more realistic road network and would possibly allow for more accuracy within cutblocks (where fine micro-topography, water tables, etc. will play a role in road development). The downside being that it would also increase computing time and may not always be necessary depending on the application of the model, especially when using the model for a large-scale landscape. Adding too much complexity to the cost surface may also take away from the goal of this model to be simple and easy to use, however, the algorithm itself would work on a very fine and detailed cost surface. Along with increased computing time, a key limitation of using this model on a finer cost surface would be that eventually cost pixels would need to have directionality associated with them, which is something this model does not take into account. The accuracy of comparing the projections to the observed network could also simply be improved by using higher quality/more up to date road and cutblock data.

Assessing the performance of this model also presented some unique challenges with comparing two networks with each other across a landscape. We needed to carefully think about what we were trying to achieve. The goal wasn't to present a model that matched the observed road network pixel by pixel, but instead was to simulate disturbance across a large-scale landscape to an acceptable level while still being a relatively simple and straightforward tool. It's for this reason why metrics were chosen that are commonly used in road impact assessments on wildlife; such as the road density and forestry disturbance footprint metrics. Once the vector networks were converted into these raster layers, it also facilitated comparing observed versus projected as we could compare the means across the TSA and study where pixels matched and didn't match. Incorporating these metrics into landscape level simulations is also a realistic goal for this tool (ie projecting forest road density across a landscape) so a goal was to therefore see how well the model did at simulating these metrics when compared to the observed landscape.

Many spatial stochastic models don't include road projections at all in their simulations (Bergeron et al., 2017; DeLong & Tanner, 1996; Naderializadeh & Crowe, 2020; Perera & Cui, 2010). Our model could be relatively easily incorporated, or applied retroactively, to similar spatial stochastic simulations such as these. For large-scale projects such as the Western Boreal Initiative (Environment Canada, date?) that simulate many complex factors across a landscape, our forest road projection methods could be incorporated as one element within the overall landscape simulations.

-The road presence metric is sensitive to small differences in road location while the road density metric is sensitive to variations in the curves of tightly packed roads within grid cells that cannot be modeled using this algorithm.

# Conclusion

The forest road network projection model presented here provides an effective way to project forest road networks that are suitable for large scale landscape simulations with relatively little effort. This model, however, is not suitable for strategic planning. The results from our projections in our study area overall matched up adequately with the observed present day forest road networks, with major discrepancies being attributed to issues in the available road and cutblock data. These discrepancies would not be issues in future landscape simulations, only when assessing accuracy. In areas where the quality of the data was high, projections matched well with the observed network. Our metrics were able to be projected effectively across a large landscape and could be useful indicators of impacts on wildlife habitats in future simulations.

This model was developed with the goal of being simple to use, readily accessible, and relatively accurate across a large landscape. With minimal input parameters, complexity, and computational power, one can project future forest road networks using this tool across large areas which can aid in the assessment of potential impacts of forestry disturbance on wildlife and habitat over a large spatial scale. With land use activities transforming a large proportion of the planet's land surface and becoming an issue of global importance (Foley et al., 2005), there is an increasing need to improve and build upon landscape level models in order to properly manage and protect remaining natural habitats. Land-use activities, primarily for agricultural expansion and timber extraction, have caused a net loss of \~7 to 11 million km2 of forest in the past 300 years (Foley et al., 2005). With only 7% of the planet's roadless areas being larger than 100 square kilometers, there is a need to protect key areas which sustain key refugia for biodiversity from fragmentation and degradation (Ibisch et al., 2016). The introduction of this widely applicable and user-friendly forest resource road model is a step in the right direction for providing more tools to land managers and planners who are interested in assessing the potential impacts of road development over large landscapes.
-->
```
# Acknowledgements

@@

## Road network projection method details

Various optimization and simulation techniques have been used for projecting forestry roads into the future and across landscapes, however, we focused on the simulation methods that link multiple harvest units (a.k.a cutblocks) to an existing road network for strategic purposes. Dean (1997) described this road projection problem as the multiple target access problem (MTAP), where the objective is to find the minimum-cost locations for new roadways that link forested stands or landings (i.e., log deck) to the existing road network. We implemented four methods for solving the MTAP in the roads package, representing a gradient of computational complexity and operational detail, starting with the least complicated or the snapping approach, followed by independent and dynamic least cost paths (LCP, DLCP) and finally, the minimum spanning tree approach (MST).

The LCP and DLCP approaches solve the MTAP by decomposing the problem into several smaller problems called single target access problems (STAP). For each STAP, a least cost path from a single source to sink is solved using Dijkstra's shortest path algorithm (1959) which minimizes the distance between a landing and the current road network, subject to costs or penalties incurred from road construction and travel. A mathematical weighted graph consisting of nodes and edges is used to parameterize Dijkstra's algorithm wherein each node corresponds to a raster pixel which may be any one of three attributes in the forest road network: (i) a landing, (ii) a point on an existing road, or (iii) an intermediate point that is a plausible location for road construction. Each node is connected with up to eight neighbours, and the cost of construction along each edge is weighted by the distance to the neighbouring node and the average of the construction cost at the source and destination nodes (or pixels) (Table \@ref(tab:cost-tbl)). The underlying grid pattern in these methods does not support curvature or switchbacks required in operational road design.

The LCP and DLCP approaches solve each STAP sequentially, starting with target that is nearest to an existing road. In the DLCP algorithm, the existing road network and construction cost surface is updated with each new road segment before moving on to the next nearest target, whereas the LCP algorithm builds a connection to each target independently. The DLCP method was first developed by Anderson and Nelson (2004) and provides a feasible solution to the MTAP but the projected network is sensitive to the ordering of the STAP. They recommended a heuristic for connecting the landings and finding optimal branch locations to deal with this sensitivity.

The MST approach extends the LCP approach by providing a heuristic for linking landings or evaluating road branching opportunities. Dean (1997) showed that heuristics with branch evaluation were able to reproduce 80% of the observed road locations in their study area. Our MST approach consists of three steps for evaluating branch locations while projecting roads: 1) estimate the least cost paths that link each of the landings and the least cost paths that link each landing to the existing road network, 2) create a weighted graph where each node is either a landing or a point on the existing road network and edge weights correspond to the total cost of the least cost path between pairs of nodes from the previous step, and 3) solve the minimum spanning tree for this graph. The minimum spanning tree is solved using Kruskal's algorithm (1956) where every node in the graph is linked to form a tree such that the sum of edge weights is minimized, and the links do not form a loop. Clark et al. (2000) also used a minimum spanning tree to project simple road networks during harvest scheduling, but they used Euclidean distance rather than least cost paths for edge weights.

TO DO: add references to appendix.

>>>>

---
title: "Implementation and assessment of methods for projecting forest road network development"
author:
  - name: Josie Hughes
    email: josie.hughes@ec.gc.ca
    institute: [ECCC]
    correspondence: true
    equal_contributor: "yes"
    orcid_id: 0000-0001-7875-9015
  - name: Sarah Endicott
    email: sarah.endicott@ec.gc.ca
    institute: [ECCC]
    correspondence: false
    equal_contributor: "yes"
  - name: David Lapins
    email: david.lapins@pc.gc.ca
    institute: [ECCC]
    correspondence: false
  - name: Kyle Lochhead
    email: Kyle.Lochhead@gov.bc.ca
    institute: [FLNR]
    correspondence: false
  - name: Gregory Paradis
    email: gregory.paradis@ubc.ca
    institute: [UBC]
    correspondence: false

institute:
  - ECCC: Environment and Climate Change Canada 
  - FLNR: British Columbia Ministry of Forests 
  - UBC: University of British Columbia 
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      number_sections: no
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: Roads.bib
csl: "../templates/journal-of-forest-ecology-and-management.csl" # Insert path for the bib-style
abstract: |
  Text of abstract
keywords: |
  keyword 1; keyword 2; keyword 3
editor_options: 
  chunk_output_type: inline
---

@@

<!-- TO DO CJFR guidelines to note:
* 6000 words, 50 references - need to remove ~30 refs
* abstract 200 words
* page #s
* double spacing
* affiliation - insitution, city, prov, country
* CMJE definition of authorship - need CRedit author contribution statement
* need funding statement, data availability statement, competing interests
* review figure guidance
-->

@@

Leston et al. (2020) projected forest road network development using Patchworks [@spatial_planning_systems_patchworks_nodate], then added fire. This example highlights that roads are also important for harvest planning, and that road network planning methods [@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997; @epstein_combinatorial_2006; @meignan_heuristic_2012] and tools exist [e.g. @remsoft_remsoft_2019; @spatial_planning_systems_patchworks_nodate]. However, algorithms contained in proprietary closed-source software are difficult to assess or fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017]. Promising options for integrating forest road network projections into large scale stochastic projections of forest change include an iterative least cost path method developed by Hardy et al [-@hardy_forest_2019; -@hardy_landis-ii_2023], and least cost path and minimum spanning tree algorithms used to project roads and reconstruct development history in British Columbia [@lochhead_demo_2018; @lochhead_linking_2022]. The former method is available as QGIS and Landis-II extensions, and the latter is embedded in a larger open source modelling project that is specific to British Columbia [@muhly_castor_nodate]. Hardy's tool is designed for integration with other components of the Landis-II modelling framework [@scheller_design_2007; @labadie_global_2023], but none of the existing implementations can be easily integrated into the full variety of modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g. @barros_empowering_2023; @daniel_state-and-transition_2016].

@@

@@

We chose the Revelstoke timber supply area (TSA 27) in British Columbia, Canada as an assessment area because it is mountainous (Figure \@ref(fig:study-area)), containing approximately 38% non-forested land where road building is difficult or impossible [new tsa data package ref]. Thus, the area provides opportunities to observe the behaviour of road projection algorithms in steep complex terrain. Forestry is the dominant resource extraction activity and much of the road network was built to access forest harvest cutblocks, allowing projection of the road network from historic forest harvest data. The Revelstoke TSA includes portions of the Monashee and Selkirk mountain ranges, and is bisected by the Columbia River. At low to mid-elevations, the forests of the interior cedar-hemlock biogeoclimatic zone (bec ref) include remnants of rare old growth inland temperate rainforest (parks canada website). At higher elevations, forests are dominated by Englemann spruce and subalpine fir (bec ref). The area is within the ancestral territories of the Ktunaxa Nation, Secwepemc Nation, Syilx and Sinixt (new tsa data package ref). Two of six

Revelstoke is in the interior-wet belt of the province where forests are dominated by Englemann spruce (31%), Hemlock (23%), Cedar (22%), and Douglas-fir (18%) [@ministry_of_forests_revelstoke_2021]. The area is home to wildlife of concern including including grizzly bears and Southern Mountain woodland caribou [@hachey_revelstoke_2010]. Eight herds of Southern Mountain caribou have been extirpated or likely extirpated, and remaining herds are considered Endangered by COSEWIC and listed as threatened under the Canadian Species at Risk Act [@palm_long_2020]. Mountain caribou avoid roads leading to a loss of available habitat [@polfus_identifying_2011; @apps_factors_2006], suffer increased mortality near roads [@apps_spatial_2013], and decline in abundance as the footprint of roads and cutblocks increases [@lochhead_linking_2022]. Roads also alter grizzly habitat use, home range selection, movements, population fragmentation, survival, reproductive rates, population density, trends, and conservation status [@proctor_effects_2020].

<!--	TO DO: consider these comments from Kyle: Need a description about road development. In BC, forestry roads are developed by forest licensees under a timber harvesting permit. Various considerations go into road development including soils, grade, and economics. What type/classification of roads are you looking at (primitive roads, paved roads, etc). What’s the maximum grade? ---can find a lot of this verbiage in the Interior Appraisal Manual. How are they maintained – do they last forever -see Auditor General Report 
•	How much of the Revelstoke TSA overlaps with caribou range? Elevation? Climate? Flora/Fauna?
•	Maybe add a table of descriptive statistics presenting information about roads in the Revelstoke TSA and the other TSAs you mention? By Caribou Range? – road density, road length, cutblock area, cutblock density, etc

-->

@@

Road network data was obtained by combining the digital road atlas [@data_bc_digital_2021] and forest tenure road [@british_columbia_ministry_of_forests_lands_natural_resource_operations_and_rural_development__flnrord_forest_2021] data sets for British Columbia. To avoid situations where the same road was included in both data sets with slightly different coordinates, we removed all roads digital road atlas that were within 50 m of forest tenure roads. The existing road network, used as the starting point for projections, was created by filtering out forest tenure roads with an award date greater than 1990. We assumed everything in the digital road atlas was built before 1990. Cutblock information was obtained from the Harvest Areas of BC (Consolidated Cutblocks) dataset [@government_of_british_columbia_harvested_2021], compiled from provincial Forest Cover, the RESULTS Reporting system, Forest Tenure applications, and satellite imagery using change detection processes. The data does not exclude non-harvested 'reserved' areas, and satellite image change detection may not always represent the entire cutblock area [@government_of_british_columbia_harvested_2021]. Forest harvest cutblocks with harvest years greater than 1990 were filtered and used as targets for the projections.

Following the Interior Appraisal Manual of BC, we assume the base cost of construction per kilometer of road is a linear function of slope [@government_of_british_columbia_interior_2021], and that road construction across major water bodies is prohibitively expensive (Table \@ref(tab:cost-tbl)). Existing roads are assigned a cost of 0. The road construction cost raster consists of 1,703,936 pixels with a resolution of 1 ha.

@@

All data sets were accessed using the `bcdata` R package and filtered to the extent of the Revelstoke Timber Supply Area [@teucher_bcdata_2021]. These data sets are continuously updated so we have stored the versions used in our analysis in an Open Science Framework repository <https://osf.io/8vmqh/>. We were not granted permission to share the digital road atlas so it must be accessed from <https://catalogue.data.gov.bc.ca/dataset/bb060417-b6e6-4548-b837-f9060d94743e>. Scripts used for data acquisition and analysis associated with this paper are available on GitHub <https://github.com/LandSciTech/RoadPaper>.

@@

We focus on methods that link multiple target locations to an existing road network for strategic purposes by finding minimum-cost locations for new roadways that link landings within harvest cutblocks (i.e., log decks) to the existing road network, described by Dean [-@dean_finding_1997] as the multiple target access problem (MTAP). Specifically, we compare dynamic least cost path (DLCP) and minimum spanning tree approaches (MST) for solving the MTAP to an alternative iterative least cost path method developed by Hardy [-@hardy_forest_2019;-@hardy_landis-ii_2023] (Figure \@ref(fig:proj-meth-fig)).

The DLCP approach solves the MTAP by decomposing the problem into several smaller single target access problems. For each landing target, the least cost path to the existing road network is solved using Dijkstra's shortest path algorithm [-@dijkstra_note_1959]. The algorithm is parameterized with a weighted graph wherein each node corresponds to a raster pixel representing one of three components of the potential forest road network: a landing, a point on an existing road, or an intermediate point that is a plausible location for road construction. Each node is connected with up to eight neighbours, and the cost of construction along edges is weighted by the distance to neighbouring nodes and the average of the construction cost at the source and destination nodes (Table \@ref(tab:cost-tbl)). The underlying grid pattern in these methods does not support curvature or switchbacks required in operational road design. The algorithm starts with the landing nearest to an existing road, and updates the existing road network and construction cost surface with each new road segment before moving on to the next nearest landing. The method was first developed by Anderson and Nelson [-@anderson_projecting_2004] who noted that the projected network is sensitive to the ordering of targets and recommended a heuristic for connecting the targets and finding optimal branch locations to deal with this sensitivity.

The MST approach extends the LCP approach by providing a heuristic for linking landings and evaluating road branching opportunities. Dean [-@dean_finding_1997] showed that heuristics with branch evaluation were able to reproduce 80% of the observed road locations in their study area. The first step in our MST approach is to estimate the least cost paths that link each of the landings and the least cost paths that link each landing to the existing road network. These results are used to create a weighted graph wherein nodes are either landings or points on the existing road network, and edge weights correspond to the total cost of the least cost path between pairs of nodes. The minimum spanning tree for this graph is solved using Kruskal's algorithm [-@kruskal_shortest_1956], which links every node with a tree that minimizes edge weights and does not include loops. This method differs from Clark et al. [@clark_three-stage_2000] who used Euclidean distances for edge weights.

The DLCP and MST methods connect landings within each cutblock to the existing road network, so an important first step is to select one or more landing locations in each block. We investigated the sensitivity of results to the pattern and density of landing locations by comparing results derived from the centroid of each cutblock to regular or randomly selected landing densities of 1 (low) and 10 (high) points per km^2^ (Figure \@ref(fig:proj-meth-fig)). We used default settings in the Hardy QGIS plugin except for skidding distance which was set to 200m. A skidding distance of 200 m corresponds to \~ 9 pts/km^2^. We compared two resolutions of the cost surface by aggregating the 1 ha resolution cost raster to a coarser 100 ha (1000 by 1000 m) resolution. In total, we compared 12 road projection method variants to a cutblocks only scenario with no projected roads (Table \@ref(tab:method-tbl)).

@@

<!-- TO DO: Sarah are these statements still correct? Please revise as necessary.-->

@@

<!-- TO DO: Sarah figure out why `r dlcp_time` and `r mst_time` are coming out as NA, and how to fix that.-->

@@

```{=html}
<!-- The model will simply build the roads in the least costly locations based on the cost surface, however often in the real world, roads must be planned around ownership and administrative boundaries. In BC, forestry roads are developed by forest licensees under a timber harvesting permit (Government of British Columbia, 2021c). -->

@@

<!-- TO DO: provide example script that would make it relatively easy for someone to assess the performance of a new road metric of their choice. They don't need to run the projections to do this - just summarize the outputs using a method of their choice. 
<!--  What David had

Mapping of roads is often especially lacking in more rural, forested, and mountainous regions where roads are often unpaved (Prendes et al., 2017) with the Auditor General of British Columbia (2017) suggesting that 100 000 km of resource roads are missing from publicly-available road network maps (inspired by Adam Ford's doc).

The model will simply build the roads in the least costly locations based on the cost surface, however often in the real world, roads must be planned around ownership and administrative boundaries. In BC, forestry roads are developed by forest licensees under a timber harvesting permit (Government of British Columbia, 2021c).

Many spatial stochastic models don't include road projections at all in their simulations (Bergeron et al., 2017; DeLong & Tanner, 1996; Naderializadeh & Crowe, 2020; Perera & Cui, 2010). Our model could be relatively easily incorporated, or applied retroactively, to similar spatial stochastic simulations such as these. For large-scale projects such as the Western Boreal Initiative (Environment Canada, date?) that simulate many complex factors across a landscape, our forest road projection methods could be incorporated as one element within the overall landscape simulations.
-->
```
# Acknowledgements

@@
